---
title: "Activity 10"
author: "Kate Miller"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include = FALSE}
# This code chunk was included to increase image quality, as well as make code chunks not visible to the reader until the code appendix. 
knitr::opts_chunk$set(echo = FALSE, dpi = 300)
```

I was unable to produce an html_document and pdf_document at the same time in the YAML heading due to an error in kableExtra. Please keep this in mind when evaluating this document.

```{r loadPackages, include = FALSE}
# This code chunk was included to load in the packages needed for future code chunks on this RMD file, using the groundhog package. 
# Loads packages with groundhog to improve stability.
library(groundhog)
pkgs <- c('ggplot2', 'tidyverse', 'tinytex', 'janitor', 'knitr', 'kableExtra', 'dplyr', 'tidyr')
groundhog.library(pkgs, '2023-11-27') 
#Use the date that you started the project here
```

## 1. Collatz Conjecture Explained

The Collatz Conjecture is a formula that begins with a starting number and counts the number of times the Collatz Conjecture has to be called before it reaches 1. If the number is even, the function runs a specific command, and if the number is odd, the function runs a different specific command.

For example, the three commands are: (1) If the number is even, the function runs number divided by 2. (2) If the number is odd, the function runs three times the number plus 1. (3) If the number is one, stop running the function.

The mathematical formula for the Collatz Conjecture is as follows: f(n) = n/2 if the number is even, f(n) = 3n + 1 if the number is odd, and f(n) = 1, STOP.

We are curious about the stopping numbers for the first 10,000 integers.

```{r collatzConjectureFunction}
# This code chunk was included to create the get_collatz_conjecture function, which takes the input of a number and outputs the number of steps. The step is determined by how many times the get_collatz_conjecture has to run or recurse until the input number reaches 1.

get_collatz_conjecture <- function(num, STEP = 0){
  # Step originates at 0, showing that the function has run 0 times initially.
  if(num == 1){
    return(STEP)
  # If 1 were the number, the step would be 0 (function does not run through at all) 
  # since no other commands would run.
  } else if(num %% 2 == 0){
    get_collatz_conjecture(num <- num/2, STEP = STEP + 1)
  # Checks to see if the number is even. 
  # If it is, the number is divided by 2 and step increases by 1.
  } else{
    num <- ((3*num) + 1)
    get_collatz_conjecture(num, STEP = STEP + 1)
  # Checks to see if the number is odd. If it is, the number is multiplied by 3 and then 
  # increased by 1, and then step increases by 1.
  }
}
```

```{r collatzConjectureVectorization}
# This code chunk was included to vectorize the get_collatz_conjecture function to convert from scalar to vector. This allows for the creation of a new object, called the collatz_conjecture, to cycle through the first 10,000 numbers (integers) through the get_collatz_conjecture function. The function returns the number of times the function had to recurse before each number reaches 1. 

vectorized_collatz_conjecture <- Vectorize(FUN = get_collatz_conjecture, vectorize.args = 'num')
# Turns the Collatz Conjecture function into a vector.
collatz_conjecture <- vectorized_collatz_conjecture(seq(from = 1, to = 10000))
# Creates an object that stores the values of the Collatz Conjecture function
# (their number of steps) from 1 to 10,000.
```

```{r collatzConjectureHistogram}
# This code chunk was included to transform the collatz_data object into a data frame that could be used to create a histogram for the first 10,000 stopping numbers in the Collatz Conjecture with their stopping times.

#| fig.cap = "Histogram of First 10,000 Digits Through Collatz Conjecture",
#| fig.pos = "H",
#| fig.width = 5 

collatz_data <- data.frame(collatz_conjecture_stop = collatz_conjecture)
# This creates a data frame for the vectorized function listed in the previous code chunk. 
ggplot(collatz_data, aes(x = collatz_conjecture_stop)) +
  geom_histogram(binwidth = 5, fill = "gray", color = "black") +
  labs(title = "Histogram of Stopping Numbers in Collatz Conjecture for 10,000 Digits",
  x = "Stopping Number", y = "Frequency") +
  theme_minimal()
# This creates a histogram of the first 10,000 numbers'stopping numbers using ggplot. 
# The axes labels are Stopping Number and Frequency. 
```

The distribution displays the number of times the Collatz Conjecture function must run for the first 10,000 numbers. It displays the frequency of each stopping number for the first 10,000 numbers. There are two peaks in the data; one around 50 and one around the 125-150 range. This means that with those two stopping numbers, there was a high frequency of numbers with those as its stopping number. In particular, the greatest frequency was around the stopping number of 50. The data dwindles down at around 200 as the stopping number, and there are very low frequencies of numbers that had a stopping number of more than 200.

## 2. Diamonds Data

The diamonds data comes from a package in R called ggplot2. The diamonds data has the attributes of price, carat, cut, color, clarity, x, y, z, depth, and table. An individual case in the data is represented by a single diamond. There are 53,940 rows and 10 columns in a data frame of the data, meaning that the attributes of 53,940 diamonds are described. The question that we are analyzing is what may contribute to the price of a diamond.

```{r diamondsDataVisualizationCaratPrice}
# This code chunk was included to create a scatter plot of the diamonds data by showing the relationship between diamond carat, price, and color.

ggplot(diamonds) +
  # Utilizes ggplot in ggplot2 to create a data visualization for price and carat of diamonds.
  aes(x = carat, y = price, colour = color) +
  geom_point(shape = "circle", size = 1.5) +
  # Specifies that we want a point data visualization (scatter plot)
  scale_color_manual(
    values = c(D = "#F8766D",
               E = "#155CA5",
               F = "#31B425",
               G = "#EAE305",
               H = "#20AFEC",
               I = "#8552D0",
               J = "#FF61C3")
  ) +
  # Selects the colors for each of the different types of diamonds.
  labs(
    x = "Carat (Diamond Weight)",
    y = "Price (USD)",
    title = "Price of Diamond Versus Carat",
    color = "Color of Diamond"
  # Details the title and the axes and legend labels. 
  ) +
  expand_limits(y = seq(0,20000), x = seq(0,5,.25)) +
  # Expands the data visualization to picture all values that are present.
  theme_minimal()
```

From this first data visualization, we can see that there is a general positive correlation between weight (carat) and price, meaning that as the weight of the diamond gets higher, the price usually also gets higher. Heavier diamonds tend to be more expensive. Color also does not seemingly affect the price.

```{r diamondsDataVisualizationCutPrice2}
# This code chunk was included to create a jitter plot of the diamonds data by showing the relationship between diamond cut, price, and color.

ggplot(diamonds) +
  aes(x = cut, y = price, colour = color) +
  # The data visualization will be between cut and price, 
  # with the points colored as colors of diamonds.
  geom_jitter(size = 1.5) +
  # Creates a jitter data visualization to see the 
  # overlapping and density of the data points for each type of cut. 
  scale_color_manual(
    values = c(D = "#CD1509",
    E = "#E86806",
    F = "#31B425",
    G = "#8080D9",
    H = "#20AFEC",
    I = "#7135CA",
    J = "#DE48A5")
  ) +
  labs(
    x = "Diamond Cut",
    y = "Price (USD)",
    title = "Price of Diamond Versus Cut",
    color = "Color of Diamond"
    # Details the title, as well as the axes and legend labels.
  ) +
  theme_minimal()

```

From this second data visualization, we can see that the the diamonds with a higher level of cut, such as premium and ideal, often reached the higher prices depicted on the graph. Some of the lower level cuts, such as fair, good, and very good, also have numerous higher-level priced diamonds. However, it is clear in the denser concentration of the premium and ideal diamonds that are above \$10,000 that, generically speaking, diamonds with a high quality cut are likely to be more expensive. Color also does not seemingly affect the price.

```{r diamondsDataSummaryTable}
# This code chunk was included to create a summary table for the diamonds data grouped by cut. The summary table displays a 7 number summary, the count, the arithmetic mean, and the arithmetic standard deviation of the diamonds data grouped by cut.

diamonds_summary_table <- diamonds %>%
  group_by(cut)%>%
  summarize(
  count = n(),
  minimum = min(price),
  first_quintile = quantile(price, 0.2),
  second_quintile = quantile(price, 0.4),
  median = median(price),
  third_quintile = quantile(price, 0.6),
  fourth_quintile = quantile(price, 0.8),
  maximum = max(price),
  mean = mean(price),
  standard_deviation = sd(price)
    )
  # The above code gathers the summary statistics for the price of 
  # the diamonds and puts it into a summary table.
diamonds_summary_table2 <- diamonds_summary_table %>%
  mutate(
    across(where(is.numeric), format, big.mark = ",")
  )
  # If the number in the table is greater than 3 digits, and is numeric, 
  # the mutate function will add commas in necessary places.

diamonds_summary_table2 %>%
  kable(
  # Utilizes the kable function to rename the 
  # columns to describe the summary statistics.
  col.names = c("Cut","Count", "Minimum", "First Quintile", "Second Quintile", "Median", "Third Quintile", "Fourth Quintile", "Maximum", "Arithmetic Mean", "Arithmetic Standard Deviation"),
  caption = "Summary Statistics of Price of Diamonds by Cut",
  booktabs = TRUE,
  align = c("l", rep("c", 6))
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("HOLD_position", 'scale_down'),
    bootstrap_options = c('striped', 'condensed'),
    font_size = 25
  # Utilizes kableExtra package to format the table in a nicely formatted way.
  ) 

```

The summary table above describes the summary statistics of the diamonds data, which includes 4 quantiles, an arithmetic mean, and an arithmetic standard deviation of the prices of the diamonds in the data set. The diamonds with the premium cut had the highest arithmetic mean price, meaning on average, they had a price of 4,584 dollars, which was higher than the other cuts. Premium diamonds also had the highest maximum price, even higher than ideal. The lowest arithmetic mean belonged to the ideally cut diamonds, with an average price of 3,457 dollars.  

## 3. Class Status

I feel that I have learned a lot in the course so far. For example, I have learned how to write functions in R (seen with the Collatz Conjecture function), how to properly name functions and objects, and how to properly format code in R. I have also learned how to import different data sets through different means (i.e. through built in packages in R and through outside Google Sheets files). Some data sets I have used in the assignments to develop my skills in R include the Army Marital Data and the Diamonds data set.

In addition, I have learned how to create tidy data frames and tidy data tables using R as well (using the two data sets aforementioned) and how to utilize the built in packages of janitor, knitr, and ggplot2 (just to name a few) to create data visualizations. Currently, I am learning how to incorporate my R coding skills, as well as documentation skills, into an R Markdown document.

``` {r ScreenshotFilePath}
# This code chunk was included to provide the code that would insert a screenshot of the file path into the RMD file.

knitr::include_graphics('file name')
```

# Code Appendix
``` {r codeAppedix}
# This code chunk was included to provide the reader with an understanding of the code behind the data visualizations depicted in the R Markdown file. It combines the code chunks together to create a code appendix.
#| ref.label = knitr::all_labels(),
#| echo = TRUE,
#| eval = FALSE
```